A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-22 inst/manuscript/manuscript.tex
minor updates before submission

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval length and coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units. This subset of population units is called a sample. In the
design-based approach, inferences about the underlying population are
informed via a probabilistic process that randomly assigns some
population units to be in the sample. Alternatively, in the model-based
approach, inferences are made from specific assumptions about the
underlying process generating the data. Each paradigm has a deep
historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the sampling or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009) and
Cicchitelli and Montanari (2012), and see Chan-Golston et al. (2020) for
a Bayesian approach).

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches that incorporate spatial
information and model-based approaches. In this manuscript, we compare
design-based approaches that incorporate spatial information to
model-based approaches for finite population spatial data. A finite
population contains a finite number of population units (we assume the
finite number is known); an example is lakes (treated as a whole with
the lake centroid representing location) in the contiguous United
States. Though we focus on finite populations, these comparisons
generalize to infinite populations as well. An infinite population
contains an infinite number of population units; an example is locations
within a single lake.

@@

Fundamentally, the design-based approach combines the randomness of the
sampling design with the data collected via the sample to justify the
estimation and uncertainty quantification of fixed, unknown parameters
of a population (e.g., a population mean). Treating the data as fixed
and incorporating randomness through the sampling design yields
estimators having very few other assumptions. Confidence intervals for
these types of estimators are typically derived using limiting arguments
that incorporate all possible samples. Sample means, for example, are
asymptotically normal (Gaussian) by the Central Limit Theorem (under
some assumptions). If we repeatedly select samples from the population,
then 95\% of all 95\% confidence intervals constructed from a procedure
with appropriate coverage will contain the true, fixed mean. Särndal et
al. (2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement some form of random sampling. When model-based approaches
implement random sampling, the inclusion probabilities are ignored when
analyzing the data (in contrast to the design-based approach, which
relies on these inclusion probabilities to analyze the data).

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of
\emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. Prediction intervals are then
derived using assumptions of the data-generating stochastic process. If
we repeatedly generate response values from the same data-generating
stochastic process and select samples, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide thorough reviews of model-based
approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual
comparison of the design-based and model-based approaches (Ver Hoef
(2002) and Brus (2021) provide similar figures).

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield narrower confidence intervals compared to
variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with a subscript \emph{u}, let

@@

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in Ver Hoef (2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
is IRS-Design. In IRS-Design, samples are selected with the Independent
Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial
locations of the population units, thus IRS samples are not spatially
balanced. In IRS-Design, samples are analyzed using the design-based
approach via the Horvitz-Thompson mean estimator and an IRS variance
estimator that does not incorporate the spatial locations of the units
in the sample. The second sampling-analysis combination is IRS-Model,
where samples are selected with the IRS algorithm and analyzed using the
model-based approach via Restricted Maximum Likelihood (REML) estimation
(Harville, 1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).
The third sampling-analysis combination is GRTS-Design, where samples
are selected with the GRTS algorithm and analyzed using the design-based
approach via the Horvitz-Thompson mean estimator and the local
neighborhood variance estimator (which does incorporate the spatial
locations of the units in the sample). The fourth and final
sampling-analysis combination is GRTS-Model, where samples are selected
with the GRTS algorithm and analyzed using the model-based approach via
REML estimation. These sampling-analysis combinations are also provided
in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS
and GRTS samples, equal inclusion probabilities were assumed for all
population units. When IRS assumes equal inclusion probabilities for all
population units, the algorithm is equivalent to simple random sampling
(SRS).

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget) from
Equation\(~\)\ref{equation:expcov}, respectively. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance was 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

Fig. \ref{fig:rmspe_eff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:rmspe_eff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So when there is no
spatial covariance, using the GRTS algorithm or a model-based analysis
does not result in much, if any, loss in efficiency compared to
IRS-Design. When there is spatial covariance (Fig. \ref{fig:rmspe_eff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to have
the lowest rMS(P)E, followed by GRTS-Design, IRS-Model, and finally
IRS-Design, though the difference in relative rMS(P)E among GRTS-Model,
GRTS-Design, and IRS-Model is relatively small. As the strength of
spatial covariance increases, the gap in rMS(P)E between IRS-Design and
the other sampling-analysis combinations widens. Finally we note that
when there is spatial covariance, IRS-Model has a much lower rMS(P)E
than IRS-Design, suggesting that the poor design properties of IRS are
largely mitigated by the model-based analysis. These rMS(P)E conclusions
are similar to those observed in the grid location layout, so we omit a
grid location layout figure here. Tables for rMS(P)E in all 36
simulation scenarios are provided in the supporting information.

@@

Fig. \ref{fig:mse_eff} shows the relative MStdE of the four
sampling-analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative MStdE is defined as
\begin{equation*}
\frac{\text{MStdE of sampling-analysis combination}}{\text{MStdE of IRS-Design}},
\end{equation*} Many general takeaways regarding MStdE are similar to
general takeaways regarding rMS(P)E: there seems to be no benefit to
using IRS, even when there is no spatial covariance; as the strength of
spatial covariance increases, the gap in MStdE between IRS-Design and
the other sampling-analysis combinations widens; and IRS-Model
outperforms (has a much lower MStdE) IRS-Design by a large margin. These
fact that the rMS(P)E and MStdE findings are similar is not particularly
surprising because the mean bias for all sampling-analysis combinations
was nearly zero, thus rMS(P)E is driven by the standard error of the
estimators (design-based) or predictors (model-based). We do note that
between GRTS-Design and GRTS-Model, GRTS-Design had lower MStdE when
there was no spatial covariance or a medium amount of spatial covariance
(Fig. \ref{fig:mse_eff}, ``Prop DE: 0'' and ``Prop DE: 0.5'' rows) and
GRTS-Model had lower MStdE when there was a high amount of spatial
covariance (Fig. \ref{fig:mse_eff}, ``Prop DE: 0.9'' row). These MStdE
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for MStdE in all
36 simulation scenarios are provided in the supporting information.

@@

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. Though we should not generalize these results to other
samples from these data, we do mention a few findings. First, IRS-Design
has the largest standard error. Second, compared to IRS-Design and
IRS-Model, GRTS-Design and GRTS-Model are much closer to the true mean
mercury concentration (have bias closer to zero) and have much lower
standard errors (more precise intervals). Third, GRTS-Model has the
least amount of bias and the lowest standard error (most precise
interval). Finally, we note that for all sampling-analysis combinations,
the true mean mercury concentration (103.2 ng / g) is within the bounds
of the combination's 95\% interval.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. The design-based and model-based approaches have
often been compared in the literature from theoretical and analytical
perspectives, but our contribution lies in studying them in a spatial
context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-Design and GRTS-Model generally had
similar rMS(P)E as their IRS counterparts when there was no spatial
covariance and lower rMS(P)E than their IRS counterparts when there was
spatial covariance. Second, the sampling decision (IRS vs GRTS) is most
important when using a design-based analysis. Though GRTS-Model still
had lower rMS(P)E than IRS-Model, the model-based analysis mitigated
most of the rMS(P)E inefficiencies that result from the IRS samples
lacking spatial balance. Third, as the strength of spatial covariance
increases, the gap in rMS(P)E and MStdE between IRS-Design and the other
sampling-analysis combinations also increases, likely because IRS-Design
is the only combination that ignores spatial locations in sampling and
analysis. Fourth and finally, when the response was normal, interval
coverage for all sampling-analysis combinations was usually close to
95\% for all sample sizes; when the response was lognormal, interval
coverage for all sampling-analysis combinations was usually between 90\%
and 95\% and closest to 95\% when \(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
also more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and response variable. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the dependence structure in the stochastic process of study. Model
selection is also possible using model-based approaches and criteria
such as cross validation, likelihood ratio tests, or AIC (Akaike, 1974).
Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Model-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations like smoothed maps. In
short, when deciding whether the design-based or model-based approach is
more appropriate to implement, the benefits and drawbacks of each
approach should be considered alongside the particular goals of the
study.

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units -- this subset is called a sample. There are two general
approaches for using samples to make frequentist statistical inferences
about a population: design-based and model-based. In the design-based
approach, inference relies on randomly assigning some population units
to be in the sample (e.g., random sampling). Alternatively, in the
model-based approach, inference relies on distributional assumptions
about the underlying stochastic process that generated the sample. Each
paradigm has a deep historical context (Sterba, 2009) and its own set of
benefits and drawbacks (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the sampling or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design-based and model-based principles (see e.g., Sterba (2009)
and Cicchitelli and Montanari (2012), and see Chan-Golston et al. (2020)
for a Bayesian approach).

Certainly comparisons between design-based and model-based approaches
have been studied in spatial contexts. But no numerical comparison has
been made between design-based approaches that incorporate spatial
locations into sampling and analysis and model-based approaches. In this
manuscript, we compare design-based approaches that incorporate spatial
locations into sampling and analysis to model-based approaches for
finite population spatial data. A finite population contains a finite
number of population units (we assume the finite number is known); an
example is lakes (treated as a whole with the lake centroid representing
location) in the contiguous United States. Though here we focus on
finite populations, the comparisons we discuss generalize to infinite
populations as well. An infinite population contains an infinite number
of population units; an example is locations within a single lake.

@@

Fundamentally, the design-based approach combines the randomness of the
sampling design with the data collected via the sample to justify the
estimation and uncertainty quantification of fixed, unknown parameters
of a population (e.g., a population mean). Treating the data as fixed
and incorporating randomness through the sampling design yields
estimators having very few other assumptions. Confidence intervals for
these types of estimators are typically derived using limiting arguments
that incorporate all possible samples. Sample means, for example, are
asymptotically normal (Gaussian) by the Central Limit Theorem (under
some assumptions). If we repeatedly select samples from the population,
then 95\% of all 95\% confidence intervals constructed from a procedure
with appropriate coverage will contain the true fixed population mean.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

The model-based approach assumes the sample is a random realization of a
data-generating stochastic process. Randomness is formally incorporated
through distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement some form of random sampling. When model-based approaches
implement random sampling, the inclusion probabilities are ignored when
analyzing the sample (in contrast to the design-based approach, which
relies on these inclusion probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of a
fixed, unknown mean, we are the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the
data-generating stochastic process. If we repeatedly generate response
values from the same process and select samples, then 95\% of all 95\%
prediction intervals constructed from a procedure with appropriate
coverage will contain their respective realized means. Cressie (1993)
and Schabenberger and Gotway (2017) provide thorough reviews of
model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we
provide a visual comparison of the design-based and model-based
approaches (Ver Hoef (2002) and Brus (2021) provide similar figures).

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield more precise (narrower) confidence intervals
compared to variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in Ver Hoef (2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.
Distributional assumptions are used, however, when constructing
prediction intervals.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
was IRS-Design. In IRS-Design, samples were selected with the
Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores
the spatial locations of the population units, thus the IRS samples were
not spatially balanced. In IRS-Design, samples were analyzed using the
design-based approach via the Horvitz-Thompson mean estimator and an IRS
variance estimator that ignored the spatial locations of the units in
the sample. The second sampling-analysis combination was IRS-Model,
where samples were selected with the IRS algorithm and analyzed using
the model-based approach via Restricted Maximum Likelihood (REML)
estimation (Harville, 1977; Patterson and Thompson, 1971; Wolfinger et
al., 1994). The third sampling-analysis combination was GRTS-Design,
where samples were selected with the GRTS algorithm and analyzed using
the design-based approach via the Horvitz-Thompson mean estimator and
the local neighborhood variance estimator (which does incorporate the
spatial locations of the units in the sample). The fourth and final
sampling-analysis combination was GRTS-Model, where samples were
selected with the GRTS algorithm and analyzed using the model-based
approach via REML estimation. These sampling-analysis combinations are
also provided in Table \ref{tab:designanalysis}. Lastly we note that for
both the IRS and GRTS samples, equal inclusion probabilities were
assumed for all population units. When IRS assumes equal inclusion
probabilities for all population units, the algorithm is equivalent to
simple random sampling (SRS).

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget) from
Equation\(~\)\ref{equation:expcov}, respectively. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), chosen so that the correlation in the dependent random
error decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the lognormal
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
lognormal random variable whose total variance was 2. The lognormal
responses were used to evaluate performance of the sampling-analysis
approaches for data that were skewed (i.e., not normal).

@@

Fig. \ref{fig:rmspe_eff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:rmspe_eff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E and using the GRTS
algorithm or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design. When there is spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``Prop DE: 0.5'' and ``Prop DE:
0.9'' rows), GRTS-Model tends to have the lowest rMS(P)E, followed by
GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in
relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is
relatively small. As the strength of spatial covariance increases, the
gap in rMS(P)E between IRS-Design and the other sampling-analysis
combinations widens. Finally we note that when there is spatial
covariance, IRS-Model has a much lower rMS(P)E than IRS-Design,
suggesting that the poor design properties of IRS are largely mitigated
by the model-based analysis. These rMS(P)E conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

Fig. \ref{fig:mse_eff} shows the relative MStdE of the four
sampling-analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative MStdE is defined as
\begin{equation*}
\frac{\text{MStdE of sampling-analysis combination}}{\text{MStdE of IRS-Design}},
\end{equation*} Many general takeaways regarding MStdE are similar to
general takeaways regarding rMS(P)E: there seems to be no benefit to
using IRS, even when there is no spatial covariance; as the strength of
spatial covariance increases, the gap in MStdE between IRS-Design and
the other sampling-analysis combinations widens; and IRS-Model
outperforms IRS-Design by a noticeable margin. These fact that the
rMS(P)E and MStdE findings are similar is not particularly surprising
because the mean bias for all sampling-analysis combinations was nearly
zero, thus rMS(P)E is driven by the standard error of the estimators
(design-based) or predictors (model-based). We do note that between
GRTS-Design and GRTS-Model, GRTS-Design had lower MStdE when there was
no spatial covariance or a medium amount of spatial covariance (Fig.
\ref{fig:mse_eff}, ``Prop DE: 0'' and ``Prop DE: 0.5'' rows), and
GRTS-Model had lower MStdE when there was a high amount of spatial
covariance (Fig. \ref{fig:mse_eff}, ``Prop DE: 0.9'' row). These MStdE
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for MStdE in all
36 simulation scenarios are provided in the supporting information.

@@

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. Though we should not generalize these results to other
samples from this population, we do mention a few findings. First,
IRS-Design has the largest standard error. Second, compared to
IRS-Design and IRS-Model, GRTS-Design and GRTS-Model are much closer to
the true mean mercury concentration (have bias closer to zero) and have
much lower standard errors (more precise intervals). Third, GRTS-Model
has the least amount of bias and the lowest standard error (most precise
interval). Finally, we note that for all sampling-analysis combinations,
the true mean mercury concentration (103.2 ng / g) is within the bounds
of the combination's 95\% interval.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. While the design-based and model-based approaches
have often been compared in the literature from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-Design and GRTS-Model generally had
similar rMS(P)E as their IRS counterparts when there was no spatial
covariance and lower rMS(P)E than their IRS counterparts when there was
spatial covariance. Second, the sampling decision (IRS vs GRTS) is most
important when using a design-based analysis. Though GRTS-Model still
had lower rMS(P)E than IRS-Model, the model-based analysis mitigated
most of the rMS(P)E inefficiencies that result from the IRS samples
lacking spatial balance. Third, as the strength of spatial covariance
increases, the gap in rMS(P)E and MStdE between IRS-Design and the other
sampling-analysis combinations also increases, likely because IRS-Design
is the only combination that ignores spatial locations in sampling and
analysis. Fourth and finally, when the response was normal, interval
coverage for all sampling-analysis combinations was usually close to
95\% for all sample sizes; when the response was lognormal, interval
coverage for all sampling-analysis combinations was usually between 90\%
and 95\% and closest to 95\% when \(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
also more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the dependence structure in the stochastic process of study.
Model selection is also possible using model-based approaches and
criteria such as cross validation, likelihood ratio tests, or AIC
(Akaike, 1974). Model-based approaches are capable of more efficient
small-area estimation than design-based approaches by leveraging
distributional assumptions in areas with few observed units. Model-based
approaches can also compute unit-by-unit predictions at unobserved
locations and use them to construct informative visualizations like
smoothed maps. In short, when deciding whether the design-based or
model-based approach is more appropriate to implement, the benefits and
drawbacks of each approach should be considered alongside the particular
goals of the study.
